Status Update regarding \n Assessment of the KLGO coastal waterbird survey monitoring program

### 2019-08-12 analysis/paper/paper.Rmd
edits to catch up post Madeleine, prep for KLGO call.

---
title: "Assessment of the KLGO Coastal Waterbird Survey Monitoring Program"
author:
- name: Joel H. Reynolds^[U.S. National Park Service, Alaska Region, Natural Resources
    Science Team]
- name: Madeleine Ward^[Swarthmore College, Dept. of Mathematics and Statistics]
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    df_print: paged
    fig_caption: yes
    number_sections: yes
    toc: yes
  bookdown::word_document2:
    fig_caption: yes
    toc: yes
  word_document:
    toc: yes
    fig_caption: true
    reference_docx: NRR_Author_Template_V4.2.docx
  pdf_document:
    fig_caption: yes
    toc: yes
editor_options: 
  chunk_output_type: inline
keywords: keyword 1; keyword 2; keyword 3
link-citations: yes
bibliography: KLGOCWS_References.bib
abstract: The Coastal Waterbird Survey has been conducted, with varying levels of
  effort and in-season frequency, since 2003 [@Surdyk2018]. The area of interest has
  been divided into eight contiguous survey units; on each survey, at each unit the
  number, age, composition (sex), breeding status, and noteworthy behavioral observations
  are recorded for each species [@Anonymous2009]. Given the constraints on personnel,
  time, and budgets, it is imperative that surveys be conducted efficiently in terms
  of both choice of sampling design and choice of sampling effort. Now that KLGO has
  multiple years of experience implementing the protocol, it is timely to step back
  and assess how things are going. We review the current program's objectives and survey design,
  identify statistical issues, and recommend potential solutions. Major topics include... blah blah. The study concludes with a series of recommended tasks that the park should
  undertake to improve the coastal waterbird survey's efficiency and effectiveness.
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

@@

```
# NOTES To Madeleine
- see edits in survey_duration.R to aid interpretation by the viewer (e.g., changes in x axis tick locations, etc.) $\checkmark$
- documentation in scripts $\checkmark$
- **NEXT Priority Milestones**: note - all references to line numbers are prior to any other edits!
- finish graphs of first & last observation per year for each target species (around line 386);**ultimately** add a fitted linear trend line to each of those (we need to account for frequency of surveying, so some subtle effort things here, but can resolve that later); $\checkmark$
- graphs of species occurrence (around line 407) for each of target species; $\checkmark$
- covariate distribution graphs (lots of univariate & bivariate things; around line 534);
- plot of total # spp sighted each season across years (around line 600)

# Issues to Resolve with content (narrative, figures, analyses)
- **Joel & Madeleine** will discuss survey_dates.R commands and how to modify them so that we get a sense of when surveys were & when target species was first (or last) seen each year. E.g., separate out any filtering process stemming from survey effort levels and timing. (thinking some thing like showing open circles for survey events PRIOR to first sighting...)
- Need to think about way to winnow out target species that do not occurr w/ adequate frequency to report on. E.g., those only seen <3 or 5 years??? Look at phenology graphs and determine threshold to apply. Consider some species N-ward and S-ward migration routes may differ.
- Joel will finish script uploading AOU data csv object & using to order Target Species
- What is up with 2008 surveys? why such distinct timing?
- Think about Covariate EDA, if any...change in observers?
- Begin thinking about (Freq of Occurrence, Phenology, Abundance | target species) and then Richness - both Phenology (Day of Season) and across years.
- clean up structure of survey design section - consistent pattern in {Element Definition, Settings, Assessment}; as cleanly as possible separate space and time aspects of survey design (& effort) from spp-specific responses.

# Introduction
Since 2003, staff at Klondike Gold Rush National Historical Park (KLGO) have surveyed coastal waterbird species distribution and abundance multiple times each year from late April through late September (Coastal Waterbird Survey or ‘CWS’; [@Surdyk2018]).  This is sufficient history to warrant stepping back and assessing the monitoring program.  Such assessments are key to an adaptive monitoring program [@Lindenmayer2009], are an important and distinct phase to promote program, and institutional, learning and evolution [@Reynolds2016], and align with guidance from NPS' Inventory and Monitoring Division [@Gallo2018a, @Mitchell2018].

Assessing a monitoring program focuses on four broad topics following the major Phases of a monitoring program (ibid):  
1. _Why_ - [Problem Framing & Information Objectives] Are the management objective(s) motivating the monitoring well articulated? Do they remain relevant and important?  
2. _How_ - [Data Collection Design] Does the monitoring data collection design effectively addresses the information objectives? Given interim learning regarding system characteristics and improvements in measurement technologies, are the information goals (including accuracy and precision goals), actually feasible? If so, can the data collection be made more efficient, reducing the observation burden on staff resources?  
3. _How_ - [Data Analysis Design] Do the data analysis methods effectively address the information objectives? Do the data analysis methods properly account for impactful features of the data collection design? Are more informative analyses available given interim improvements in methods?  
4. _How_ - [Workflows & Reporting] Are workflows and related software systems in place to allow timely, and efficient, data management, analysis updating, reporting, and distribution of findings to target audiences of information users?

This assessment addressed all four topics, but focused predominantly on topics (i) Problem Framing & Information Objectives, (ii) Data Collection Design, and (iii) Data Analysis Design.  Since 'Problem Framing & Information Objectives' determine the relevancy of many of the potential assessment actions related to the other topics,  the assessment was drafted as a first stage of an iterative effort, providing high level feedback, including listings of potential detailed syntheses and workflow enhancement activities (such as improved database interface, automation of annual analyses and updating of syntheses, etc.).  With this foundation, further dialogue with KLGO staff regarding relevant 'Problem Framing & Information Objectives' and other priorities will determine which follow-on activities to pursue.

<!-- Key Findings / Next Steps -->

@@

# Methods
The CWS monitoring program was compared against a recommended checklist of monitoring program components developed from published literature (Appendix 1). It identifies multiple component decisions associated with each of the major Phases of a monitoring program (briefly overviewed above).  For each component, the relevant aspects of the current CWS were reviewed, potential issues identified, potentially informative diagnostic analyses suggested, and recommendations provided.  The characterization of the CWS was based on KLGO protocols and reports (Table 1), a site visit (6/4-6/2018) and personal communications with the KLGO Natural Resources Program Manager (Jami Belt, winter and spring 2018), KLGO Resources Manager (Anne Matsov, summer 2018) and long-standing survey participants (Elaine Furbish, Skagway Bird Club, spring and summer 2018).

@@

The management problem or question motivating the need for information about coastal waterbirds provides the foundation for the survey’s data collection and analysis designs [@Fancy2009].  It clarifies the underlying goal of the monitoring program and answers “Why is this an important problem or question?”.

@@

### Assessment findings:
The 2003 coastal waterbird inventory and survey design was developed to address "a top priority bird inventory (need)" of the Southeast Alaska Inventory and Monitoring Network (Sharman et al. 2000) - namely, an inventory of waterfowl/shorebirds at KLGO [@Hahr2004].  This was deemed a priority due to the area’s expected avifaunal richness [@NABCI2000] and the limited availability of reliable observation records - mainly consisting of systematic surveys conducted by NPS staff in the early 1980s [@Hahr2004].

The inventory's stated goal was to document the occurrence of 90% of the species of waterbirds likely to occur in KLGO (ibid). The principal objectives were “to develop scientifically valid sampling designs for inventorying waterbirds and breeding landbirds in KLGO, and to collect baseline information on bird species occurrence, distribution, abundance, and habitat associations in the park…[to] ...provide park managers with timely information that facilitates land management planning, helps track wildlife population and ecosystem change over time, and keeps resource managers apprised of changes that may require action before a crisis occurs” (ibid).  While no Threatened or Endangered bird species were known to occur in KLGO, as of 2004, nine species on the Alaska Watchlist had been documented in the park (ibid), thus suggesting potential priority information needs associated with monitoring for changes in those species occurrence and abundance.  Potential secondary objectives mentioned include "investigations into the breeding ecology, phenology, reproductive success and population age structure of KLGO breeding birds" (ibid). Later reports mention additional objectives of estimating changes in species occurrence and composition [@St.Pierre2015], characterizing timing of peak spring waterbird migration and its relation to spring eulachon runs [@Surdyk2017].

Stakeholders mentioned or implied in park reports include park superintendent and related managers, those charged with monitoring regional and internationally important bird habitats, the Skagway Bird Club, the Alaska Landbird Monitoring Survey and the Boreal Partners in Flight.

The coastal waterbird surveys have been continued since 2003 [@Surdyk2018].

QUOTES
Hahr and Trapp (2004):
We intend to use the results of these inventories to create a revised bird checklist and an atlas database for the Taiya and Skagway River watersheds in collaboration with the Skagway Bird Club.

This protocol addresses one monitoring question with three objectives: (1) what is the annual nest density and reproductive (fledgling) success in known Hawaiian petrel colonies, (2) what are the long-term trends in colony distribution and density monitored in approximate 5-year intervals, and (3) are these affected by predator control? The first goal of monitoring is to obtain unbiased estimates of Hawaiian petrel nest density and reproductive (fledging) success from known colonies in HAVO and HALE in order to detect changes in colony growth or decline. The second goal of monitoring is to periodically (approximately every five years) obtain unbiased estimates of nest density from potential Hawaiian petrel habitat. This information can be used to assess changes in density and distribution of subcolonies across the landscape. The third goal is to estimate nest density and fledging success in areas undergoing different management regimes to assess effectiveness of management. Benchmark levels of these estimates could serve as warnings of the need for modified management or further investigation of these colonies.

Though not explicitly characterized in the protocol or reports reviewed for this assessment (Table 1), the following target audiences and information users were identified through conversations with KLGO staff, partners, and other waterbird researchers and stakeholders in Alaska.
(current)
- park
- Skagway Bird Club; compiles for addition to 'significant bird observations in SE AK', which gets compiled in to AK sumamries for 'North American Birds' (mainly spp lists)
(potential)
- BPIF
- PSG
- Audubon Alaska (for IDing Important Bird Areas - though require accurate density estimates)
- Audubon North America - cliamte species distribution models (Nicole Michel, Brooke Bateman, etc.); esp if associated w/ repeat sites & effort data (date, time of day, duration, distance/area covered) - see Nicole email 6/7/2018. Watchlist of projected changes - [@Wu2018]

### Recommendations:

Potential Follow Up Activities:

Define the _Target Audiences_ for the information and their Objectives in using the information

_Decision Makers_
Who are target audiences for this information?
Who uses it or may use it?
Local contact: C. Elaine Furbish, Skagway Bird Club, Google Site
Predominantly historical park; limited resource staff and capacity
Visible natural resource of interest to local community and visitors; interp and educational opportunity; topic of information requests; provides ‘community service’ given absence of any other agency with natural resource staff;
Annual report (time lag?); National Resource Condition Assessment; other? State of the Park?
Potential for contribution to regional, flyway, and WASO programs (e.g., CCRP & Audubon climate watch, etc.)
Current sightings passed to local bird club & used by avian-interested visitors
Used in quarterly regional ‘significant bird observations’, which in turn is used to develop the Alaska-wide summary for the journal North American Birds (published by the American Birding Association http://www.aba.org/nab ) and includes ‘...sightings of birds that are out of range or out of season or occur regularly in small numbers, noteworthy breeding records, unusually large or small numbers of a particular species, unusual migration dates, etc….’ (S. C. Heinl, 2017, Summary of SE AK Bird Observations, June-July 2017, accessed 2018 June 6).
Queries to Heather Renner & Robin Corcoran re: PSG-type users, etc.
Opportunity for community based monitoring to track projected spp impacts of climate change, change in phenology, occurrence, abundance, etc.?

_Decisions_
What types of decisions is it / will it be used for?
(specify for each target audience/ end user)
No specific decisions
Current:
KLGO - spp lists, phenology of occurrence (guidance to visitors), trends in richness (NRCA), abundance?
Skagway bird club - spp list, current occurrences (activity re: rare & unusual)
SE AK / Regional - spp lists; unusual sightings, etc. (spp occurrences & broad seasonal timing).
AKR -
WASO/NPS - ??CCRP?? Question to Gregor & Mel (AK Audubon). Could be used for garnering community engagement in Climate Watch & other national efforts associated with Audubon’s Climate Report. Not used in continental analyses.
“One key message that I want to emphasize is that the focus of this work was purely on climate suitability.  What this means, as we indicate in the briefs’ “Important” box up front and “Caveats” section further back, is that significant changes in climate suitability, as measured here, will not always result in a species response, and all projections should be interpreted as potential trends. Multiple other factors mediate responses to climate change, including habitat availability, ecological processes and biotic interactions, dispersal capacity, species' evolutionary adaptive capacity, and phenotypic plasticity (e.g., behavioral adjustments). Ultimately, models can tell us where to focus our concern and which species are most likely to be affected, but monitoring is the only way to validate these projections and should inform any on-the-ground conservation action.” Gregor Schuurman, email to DENA

Summarize the management or policy decisions the monitoring is to inform (if any) based on 3.c.

## Phase 2 - Monitoring & Analysis Design
HOW monitor? Clarify the various design choices.

@@

*Site Accessibility*
How consistently are they able to conduct surveys at each site? E.g., how often is Site 1 dropped due to cruise ships?

> SECOND graph: [THIS IS SECONDARY PRIORITY] x axis is Time of Day, perhaps broken in to 1 or 2 hour blocks , y axis is ....this is a tough one so totally open to your ideas. How about y axis is % [0, 1.0] and for each year there is a different curve showing the % of planned surveys (as defined above) that were successfully completed that season within that time block; facet by site. So we are seeing whether there are consistently problematic times of day at each site. 
> JOEL: for now, times are in hour blocks

```{r SiteAccess, eval=TRUE, cache = FALSE}

source("../scripts/survey_access.R", echo=FALSE)
```

```{r Fig-SiteAccess1, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis',fig.height=20, fig.width=12, fig.cap="",warning= FALSE}

#print(plot_access)
```

@@

*Survey Completion*
> Ultimately, re-order these sections to flow better. Completion then Duration, etc., etc.

Add sentence or two about success in completing surveys. Doesn't appear to be any consistently problematic sites, not even site 1 (Figure \@ref(fig:Fig-SiteSurveyCompletion)).

**MADELEINE: Dig in to Site 1 to make sure there aren't messages hidden in 'notes' since that is the location that consists of cruise ship docking areas for a good chunk of the space.**

print(plot_unitSuccess)
```

*Temporal* ??how define ‘season’??

The annual dates of the first and last survey each year has varied by over a month (Figure \@ref(fig:Fig-SurveyTiming)).

```{r EDA_SurveyTiming, eval=TRUE, cache = FALSE, include = FALSE}
# dotplot of survey times by year

```{r Fig-SurveyTiming1, echo=FALSE, eval=TRUE, cache=FALSE, fig.show= 'asis',fig.height=8, fig.width=12, fig.cap="The annual dates of the first and last survey have both varied by over a month across the years."}

print(plot_alldates)
```

@@

##### pagebreak

>>>>

---
title: "Status Update regarding \n Assessment of the KLGO coastal waterbird survey monitoring program"
author:
- name: Joel H. Reynolds^[U.S. National Park Service, Alaska Region, Natural Resources
    Science Team]
- name: Madeleine Ward^[Swarthmore College, Dept. of Mathematics and Statistics]
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::word_document2:
    fig_caption: yes
    toc: no
  pdf_document:
    fig_caption: yes
    toc: no
  word_document:
    toc: no
    fig_caption: true
    reference_docx: NRR_Author_Template_V4.2.docx
  html_document:
    df_print: paged
    fig_caption: yes
    number_sections: yes
    toc: no
editor_options: 
  chunk_output_type: inline
keywords: keyword 1; keyword 2; keyword 3
link-citations: yes
bibliography: KLGOCWS_References.bib
abstract: DRAFT ABSTRACT The Coastal Waterbird Survey has been conducted, with varying levels of
  effort and in-season frequency, since 2003 [@Surdyk2018]. The area of interest has
  been divided into eight contiguous survey units; on each survey, at each unit the
  number, age, composition (sex), breeding status, and noteworthy behavioral observations
  are recorded for each species [@Anonymous2009]. Given the constraints on personnel,
  time, and budgets, it is imperative that surveys be conducted efficiently in terms
  of both choice of sampling design and choice of sampling effort. Now that KLGO has
  multiple years of experience implementing the protocol, it is timely to step back
  and assess how things are going. We review the current program's objectives and survey design,
  identify statistical issues, and recommend potential solutions. Major topics include... etc. The study concludes with a series of recommended tasks that the park should
  undertake to improve the coastal waterbird survey's efficiency and effectiveness.
---

@@

# Overview of Progress on this TAR
* Brief overview of activities
  + 2018 Summer - site visit KLGO; discussions w/ Jamie, Elaine Furbish; gather protocol and past report materials, database, etc. Develop assessment strategy and report outline; initial review of past reports and related materials. Recruit 10 hr / wk volunteer undergraduate student intern (Madeleine Ward).
  + 2018 Fall-Winter - bulk of effort ended up focused on working with Madeleine and NPS WASO IT staff to set up computing systems for functional collaborative workflow (e.g., establich NPS github repository and train intern in use of 'version control systems' to automate synchronizing of our files for collaboration, resolving various IT issues, setting up project folder and file structures for long-term use and archiving); initial assessment of program goals and objectives; resolve data access issues; 
  + 2019 Spring - initiate basic summary graphics for clarifying timing and frequency of survey events, variation in duration of survey events (e.g., variation in survey effort at a site); timing and frequency of spp-specific sightings; intial development of phenology graphics. INTERNSHIP ENDED
  + 2019 Summer (minimal time) - prep for discussion w/ KLGO Resources staff.

The following excerpt from the planned report lists 4 major topics or phases that should be addressed in a full assessment of a monitoring program.  The most important is the first topic - the *Why* motivating the program; e.g., the program's Problem Framing and Information Objectives. I've provided draft feedback on this and a number of key questions that require resolution by KLGO staff prior to diving into any detailed analyses *so as to ensure futher detailed review of data collection & analyses target KLGO's desired Information Objectives*. I've also included some of the basic survey effort data visualizations that we've already developed but haven't spent time on the associated narrative.

***
# Introduction
Since 2003, staff at Klondike Gold Rush National Historical Park (KLGO) have surveyed coastal waterbird species distribution and abundance multiple times each year from late April through late September (Coastal Waterbird Survey or ‘CWS’; [@Surdyk2018]).  This is sufficient history to warrant stepping back and assessing the monitoring program.  Such assessments are key to an adaptive monitoring program [@Lindenmayer2009], are an important and distinct phase to promote program, and institutional, learning and evolution [@Reynolds2016], and align with guidance from NPS' Inventory and Monitoring Division [@Gallo2018a], [@Mitchell2018].

Assessing a monitoring program focuses on four broad topics, each associated with a major development phase of the monitoring program (ibid):  
1. _Why_ - [Problem Framing & Information Objectives] Are the management objective(s) motivating the monitoring well articulated? Do they remain relevant and important?  
2. _How_ - [Data Collection Design] Does the monitoring data collection design effectively addresses the information objectives? Given interim learning regarding system characteristics and improvements in measurement technologies, are the information goals (including accuracy and precision goals), actually feasible? If so, can the data collection be made more efficient, reducing the observation burden on staff resources?  
3. _How_ - [Data Analysis Design] Do the data analysis methods effectively address the information objectives? Do the data analysis methods properly account for impactful features of the data collection design? Are more informative analyses available given interim improvements in methods?  
4. _How_ - [Workflows & Reporting] Are workflows and related software systems in place to allow timely, and efficient, data management, analysis updating, reporting, and distribution of findings to target audiences of information users?

This assessment aims to address all four topics, but will focus predominantly on topics (i) Problem Framing & Information Objectives, (ii) Data Collection Design, and (iii) Data Analysis Design.  Since 'Problem Framing & Information Objectives' determine the relevancy of many of the potential assessment actions related to the other topics,  the assessment was drafted as a first stage of an iterative effort, providing high level feedback, including listings of potential detailed syntheses and workflow enhancement activities (such as improved database interface, automation of annual analyses and updating of syntheses, etc.).  With this foundation, further dialogue with KLGO staff regarding relevant 'Problem Framing & Information Objectives' and other priorities will determine which follow-on activities to pursue. 
<!-- Key Findings / Next Steps -->

@@

# Assessment Methods
The CWS monitoring program was compared against a recommended checklist of monitoring program components developed from published literature (Appendix 1). It identifies multiple component decisions associated with each of the major Phases of a monitoring program (briefly overviewed above).  For each component, the relevant aspects of the current CWS were reviewed, potential issues identified, potentially informative diagnostic analyses suggested, and recommendations provided.  The characterization of the CWS was based on KLGO protocols and reports (Table 1), a site visit (6/4-6/2018) and personal communications with the KLGO Natural Resources Program Manager (Jami Belt, winter and spring 2018), KLGO Resources Manager (Anne Matsov, summer 2018) and long-standing survey participants (Elaine Furbish, Skagway Bird Club, spring and summer 2018).

@@

Ideally, the survey’s data collection and analysis designs are founded on clearly articulated *Information Needs* stemming from a clearly articulated management problem or question or goal  [@Fancy2009].  Such a problem or question or goal answers “Why is this information important?”.

@@

### Assessment findings:
#### Motivating Management Problem, Question or Goal
The 2003 coastal waterbird inventory and survey design was developed to address "a top priority bird inventory (need)" of the Southeast Alaska Inventory and Monitoring Network (Sharman et al. 2000) - namely, an inventory of waterfowl/shorebirds at KLGO [@Hahr2004].  This was deemed a priority due to the area’s expected avifaunal richness [@NABCI2000] and the limited availability of reliable observation records - mainly consisting of systematic surveys conducted by NPS staff in the early 1980s [@Hahr2004].

The inventory's stated goal was to document the occurrence of 90% of the species of waterbirds likely to occur in KLGO "... [to] ...provide park managers with timely information that facilitates land management planning, helps track wildlife population and ecosystem change over time, and keeps resource managers apprised of changes that may require action before a crisis occurs" (ibid).

=> Thus the program appears to be intended to serve management information needs associated with land use planning, provide 'surveillance monitoring' of population changes and trends through time, and provide red-flags for potential developing crises.

**Recommendations**

=> **Park Resource staff should clarified and further refined the motivating needs in order to allow assessment of the adequacy of the current survey design and effort levels.**

What actual land management questions might such information inform? Have any occurred in the last 20 years? If that remains a valid goal, can the types of questions and decisions be refined and clarified? What sort of crises? How quickly do they need to be 'flagged'? What sort of ability to detect trend? Only local trend or regional or...? Trend in phenology or occurrence frequency or abundance or some other population feature? Clarifying these goals matters, for example, since monitoring phenology requires much less intensive survey effort than effectively monitoring abundance.

#### Motivating Management Information Objectives
The principal objectives were identified as (numbering added) “(i) to develop scientifically valid sampling designs for inventorying waterbirds and breeding landbirds in KLGO, and to collect baseline information on bird species (ii) occurrence, (iii) distribution, (iv) abundance, and (v) habitat associations in the park ” (ibid).

While no Threatened or Endangered bird species were known to occur in KLGO, as of 2004, nine species on the Alaska Watchlist had been documented in the park (ibid), thus suggesting potential priority information needs associated with monitoring for changes in those species occurrence and abundance.

Potential secondary objectives mentioned include "investigations into the breeding ecology, phenology, reproductive success and population age structure of KLGO breeding birds" (ibid). Later reports mention additional objectives of estimating changes in species occurrence and composition [@St.Pierre2015], characterizing timing of peak spring waterbird migration and its relation to spring eulachon runs [@Surdyk2017].

=> depending on how you classify things, this is up to 12 distinct information objectives.  Distinct information objectives may well require very different surveys methods, if not at least different levels of survey effort, and analyses.

**Recommendations**

=> **Park Resource staff should revisit and evaluate the relevance and importance or each of these distinct information objectives**, ideally ranking them at least into 'essential' versus 'nice if feasible' categories. This will provide the targets for actually assessing adequacy of the current survey design and analysis methods.

#### Key Stakeholders
Stakeholders mentioned or, more usually, implied in park reports include park superintendent and related managers, those charged with monitoring regional and internationally important bird habitats, the Skagway Bird Club (author of 'significant bird observations in SE AK' and contributor to AK summaries for 'North American Birds'), the Alaska Landbird Monitoring Survey and the Boreal Partners in Flight.  For example, Hahr and Trapp (2004) mention the intention to develop inventories and revise the bird checklist and atlas database for the Taiya and Skagway River watersheds in collaboration with the Skagway Bird Club.

**Recommendations**

=> **KLGO Natural Resources staff should explicitly clarify intended stakeholders and 'information users', clarify their exact information needs in order to assure they will be met (or cannot be met), and develop list of contacts for distributing results and updates (perhaps includes as Appendix in SOP if not already).**

=> Additional potential stakeholders include Audubon North America as part of their national Climate Watch program, which develops species distribution models under projected future climates. This group is especially interested in data from areas with repeat observations at sites and that include recording of effort data (date, time of day, duration, distance/area covered, etc.).  An example of their products is the recent watchlist of projected changes [@Wu2018]. 
(Nicole Michel, Brooke Batemen; see email from Nicole to Joel 6/7/2018)

## Phase 2 - Monitoring & Analysis Design
**NOTE TO READER**: The rest of this is a mix of outline and notes, with some initial data visualizations. This is far from a draft narrative. **Proceed with Caution**

HOW monitor? Clarify the various design choices.

@@

*Temporal* 
```{r EDA_SurveyTiming, eval=FALSE, cache = FALSE, include = FALSE}
# dotplot of survey times by year

```{r Fig-SurveyTiming1, echo=FALSE, eval=FALSE, cache=FALSE, fig.show= 'asis',fig.height=8, fig.width=12, fig.cap="The annual dates of the first and last survey have both varied by over a month across the years."}

print(plot_alldates)
```
The annual dates of the first and last survey each year has varied by over a month (Figure \@ref(fig:Fig-SurveyTiming)).

*Survey Completion*

print(plot_unitSuccess)
```
Add sentence or two about success in completing surveys. Doesn't appear to be any consistently problematic sites, not even site 1 (Figure \@ref(fig:Fig-SiteSurveyCompletion)).

@@

@@

***
#### Misc draft text currently removed
_Decision Makers_
Who are target audiences for this information?
Who uses it or may use it?
Local contact: C. Elaine Furbish, Skagway Bird Club, Google Site
Predominantly historical park; limited resource staff and capacity
Visible natural resource of interest to local community and visitors; interp and educational opportunity; topic of information requests; provides ‘community service’ given absence of any other agency with natural resource staff;
Annual report (time lag?); National Resource Condition Assessment; other? State of the Park?
Potential for contribution to regional, flyway, and WASO programs (e.g., CCRP & Audubon climate watch, etc.)
Current sightings passed to local bird club & used by avian-interested visitors
Used in quarterly regional ‘significant bird observations’, which in turn is used to develop the Alaska-wide summary for the journal North American Birds (published by the American Birding Association http://www.aba.org/nab ) and includes ‘...sightings of birds that are out of range or out of season or occur regularly in small numbers, noteworthy breeding records, unusually large or small numbers of a particular species, unusual migration dates, etc….’ (S. C. Heinl, 2017, Summary of SE AK Bird Observations, June-July 2017, accessed 2018 June 6).
Queries to Heather Renner & Robin Corcoran re: PSG-type users, etc.
Opportunity for community based monitoring to track projected spp impacts of climate change, change in phenology, occurrence, abundance, etc.?

##### misc notes re: potential (Stakeholder, Info Needs) combos
_Decisions_
What types of decisions is it / will it be used for?
(specify for each target audience/ end user)
No specific decisions
Current:
KLGO - spp lists, phenology of occurrence (guidance to visitors), trends in richness (NRCA), abundance?
Skagway bird club - spp list, current occurrences (activity re: rare & unusual)
SE AK / Regional - spp lists; unusual sightings, etc. (spp occurrences & broad seasonal timing).
AKR -
WASO/NPS - ??CCRP?? Question to Gregor & Mel (AK Audubon). Could be used for garnering community engagement in Climate Watch & other national efforts associated with Audubon’s Climate Report. Not used in continental analyses.
“One key message that I want to emphasize is that the focus of this work was purely on climate suitability.  What this means, as we indicate in the briefs’ “Important” box up front and “Caveats” section further back, is that significant changes in climate suitability, as measured here, will not always result in a species response, and all projections should be interpreted as potential trends. Multiple other factors mediate responses to climate change, including habitat availability, ecological processes and biotic interactions, dispersal capacity, species' evolutionary adaptive capacity, and phenotypic plasticity (e.g., behavioral adjustments). Ultimately, models can tell us where to focus our concern and which species are most likely to be affected, but monitoring is the only way to validate these projections and should inform any on-the-ground conservation action.” Gregor Schuurman, email to DENA

#### Misc QUOTES for potential use later
This protocol addresses one monitoring question with three objectives: (1) what is the annual nest density and reproductive (fledgling) success in known Hawaiian petrel colonies, (2) what are the long-term trends in colony distribution and density monitored in approximate 5-year intervals, and (3) are these affected by predator control? The first goal of monitoring is to obtain unbiased estimates of Hawaiian petrel nest density and reproductive (fledging) success from known colonies in HAVO and HALE in order to detect changes in colony growth or decline. The second goal of monitoring is to periodically (approximately every five years) obtain unbiased estimates of nest density from potential Hawaiian petrel habitat. This information can be used to assess changes in density and distribution of subcolonies across the landscape. The third goal is to estimate nest density and fledging success in areas undergoing different management regimes to assess effectiveness of management. Benchmark levels of these estimates could serve as warnings of the need for modified management or further investigation of these colonies.

# To Do
* Issues to Resolve with content (narrative, figures, analyses)
  + Framework - review & incorporate relevant from I&M national program guidance for frequency of analyses & reporting, etc., etc.
  + Winnow out target species that do not occurr w/ adequate frequency to report on. E.g., those only seen <3 or 5 years??? Look at phenology graphs and determine threshold to apply. Consider some species N-ward and S-ward migration routes may differ.
  + finish script uploading AOU data csv object & using to order Target Species
  + What is up with 2008 surveys? why such distinct timing?
  + Think about Covariate EDA, if any...change in observers?
  + Begin thinking about (Freq of Occurrence, Phenology, Abundance | target species) and then Richness - both Phenology (Day of Season) and across years.
  + clean up structure of survey design section - consistent pattern in {Element Definition, Settings, Assessment}; as cleanly as possible separate space and time aspects of survey design (& effort) from spp-specific responses.
  + covariate distribution graphs (lots of univariate & bivariate things; around line 534);
  +plot of total # spp sighted each season across years (around line 600)
